Investigating magnetic dynamo evolution with TESS field dwarfs

### 2017-10-03 ms.tex
some changes to the text

Among the key observable properties of main sequence stars in our Galaxy,
age is the most difficult to determine.
Traditionally, fitting isochrones to cluster stars was one of the only precise
methods for measuring ages but was extremely difficult for the majority of
isolated field stars, particularly for those without precise spectroscopic
information.
Methods such as asteroseismology and measuring Lithium abundances can provide
precise ages but require time intensive observations for each target and are
not capable of producing the large quantity of ages needed for exoplanet and
galactic population studies.

Galactic archaeology and exoplanet populations are two rapidly accelerating
fields of interest within astronomy. Although seemingly unconnected, these two
fields are linked by a mutual requirement for precise stellar parameters.
To galactic archaeologists, ages and elemental abundances are the most
important parameters.
Indeed, most galactic archaeology surveys target exactly these properties.
For exoplaneteers, masses and radii have historically been the most important
stellar parameters for understanding planetary systems.
With a growing number of planet hosts with precise masses and radii, attention
is turning toward other parameters such as ages to understand the history and
evolution of these systems.
Age is therefore a fundamental stellar parameter of great interest to two
large communities of astronomers.
However it is a difficult attribute to measure for main sequence F, G, K, and
M stars in the field, in part because low-mass dwarfs do not move far on the
Color-Magnitude diagram (CMD) during their hydrogen burning lifetimes.
Further, competing stellar evolution models predict different ages for the
same star.
Of the measurable properties for a large ensemble of field stars, rotation
periods contain the most information about stellar age, and provide the best
leverage for advancing our knowledge of galactic archeology as well as
exoplanet population demographics.

To improve our understanding of star and planet formation and evolution, as
well as the history of the Milky Way, we must constrain the ages of low-mass
stars like the Sun in the galactic field.
Fortunately nature has provided a powerful means to determine ages for main
sequence stars via their rotation.
Angular momentum is carried away though magnetically driven stellar winds,
which slows the star’s rotation over cosmic time. This rotation-based “clock”
is known as gyrochronology.
Cool spots on the star’s surface rotate in-to and out-of view, creating small
amplitude ($\pm$ 1\%) quasi-periodic changes in the stellar brightness.
While rotation periods have previously been measured from starspot-induced
flux modulations for hundreds of stars from the ground, space-based
photometric surveys have opened the door to homogeneous ensemble measures of
stellar rotation, and therefore age.
With precise light curves available from the TESS mission, we may be able to
determine rotation periods and ages for nearly 100,000 main sequence field
stars.

@@

To enable studies of stellar ages from rotation periods with TESS, we propose
to: 1. Measure accurate rotation periods for every available TESS FFI and
two-minute cadence target, using the most appropriate tools and methods
required.
2. Produce updated gyrochronology relations based on a wider range of field
star ages.
3. Determine the origin of the mysterious rotation period bimodality
discovered with Kepler by tracing the rotation period distribution over the
whole sky, and out to further distances utilizing public Gaia data.
4. Measure the star formation history across the sky using a new Bayesian
age-dating system.

@@

the discovery of a bimodal period distribution among field stars by McQuillan
et al. (2013), which is shown
in Figure \ref{davenport}.

@@

The K2 rotation period sample provides the ideal dataset to test these two

@@

this feature as seen in Figure 2b with galactic position, distance, or between
K2 fields.
this feature as seen in Figure 2b with galactic position, distance, or between
K2 fields.

@@

Intriguingly, the rotation gap is not constant across masses.
It appears at shorter rotation periods for more massive stars and longer
rotation periods for less massive stars.
In fact, it closely follows a contour of constant age, an isochrone of 600
Myr.
Two possible scenarios can explain this phenomenon.
The first is that stars undergo a transition in the behaviour of their
magnetic dynamo.
They switch from a fast to a slow rotating regime rapidly, spending little
time rotating at an intermittent period.
This transition would occur around 3-4 days for G dwarfs and 10-12 days for M
dwarfs.
Is this a convergence bump?
The rotation periods of stars are believed to have converged by around the age
of the Hyades, ~790 Myr (clusters younger than this show a broad spread in
rotation periods and older clusters a narrow one).
The age at which this transition occurs, around 600 Myr, is close to this.
Could this be a coincidence or do rotation periods undergo a rotation period
bump once they move stabilise onto a Skumanich spin-down regime?
If this is the case, a study of this transition could greatly advance our
understanding of the operation of stellar dynamos.

The rotation period bimodality seen in McQuillan et al. (2014) and Davenport
(2017) could also be explained by two waves of star formation: a short one
centered around a few hundred million years ago and a long one centered around
a few billion years ago.
TESS’s full view of the sky will allow us to establish the spatial extent of
this rotation period gap and investigate the underlying cause.

@@

Gyrochronology is an active area of interest in astronomy today due to its applicability to stellar, exoplanet and galactic astronomy.
It is a method of age-dating stars that doesn’t require expensive
spectroscopy, requiring only a rotation period and a colour or temperature.
Considering the vast number of light curves that will be made available by
future photometric surveys, as well as the already formidable number produced
by the Kepler spacecraft, gyrochronology is by far the most widely applicable,
relatively precise age-dating method available.
Although the phenomenon of magnetic braking has been closely studied since its
discovery in the mid/late 20th century, the manifestations of magnetic fields
and their effects on rotation and activity are complex; the physics
underpinning rotational evolution are far from well-understood.
In fact, concerning rotation, though Kepler was a triumphant planet hunter and
a revolutionary stellar mission, it may have revealed as many new mysteries as
it resolved old ones.
One mystery is described above; it is the appearance of a bimodal period
distribution.
The other concerns gyrochronology, the method of inferring the age of a star
from its rotation period and mass.

Although the classical spin-down law of Skumanich (1972), Period $\propto$
Age$^1/2$ holds for all open clusters with measured periods, it does not
appear to describe old field stars (Angus et al., 2015, van Saders et al.,
2016).
Asteroseismic pulsators observed by Kepler, older than the Sun, rotate more
rapidly than the Skumanich law predicts they should. Van Saders et al. (2016)
identified a mass-dependence in the… recalibrated the gyrochronology
relations, accounting for this divergence from the fiducial spin-down
relation.
They attribute the difference between old and new models to weakened magnetic
braking.
Once a star reaches a critical Rossby number (the ratio of the rotation period
to the convective overturn time), its magnetic dynamo transitions into a
different state which drastically reduces (basically switches off) braking
efficiency.
This means a 5 Gyr star of Solar mass will have the same rotation period as a
6 Gyr and rotation period cannot be used as an age proxy.
The transition occurs at around Solar age for Solar mass stars but later for
less massive stars and earlier for more massive stars.
So, although this effect reduces the usefulness of gyrochronology for stars
more massive than the Sun, in practise, for Solar masses and below,
gyrochronology can still be considered an effective dating method.

>>>>

% Among the key observable properties of main sequence stars in our Galaxy,
% age is the most difficult to determine.
% Traditionally, fitting isochrones to cluster stars was one of the only precise
% methods for measuring ages but was extremely difficult for the majority of
% isolated field stars, particularly for those without precise spectroscopic
% information.
% Methods such as asteroseismology and measuring Lithium abundances can provide
% precise ages but require time intensive observations for each target and are
% not capable of producing the large quantity of ages needed for exoplanet and
% galactic population studies.

Galactic archaeology and exoplanet populations are two rapidly accelerating
fields of interest within astronomy.
Although seemingly unconnected, these fields are linked by a mutual
requirement for precise stellar parameters.
To galactic archaeologists, ages and elemental abundances are the most
important parameters.
Indeed, most galactic archaeology surveys target exactly these properties.
For exoplaneteers, masses and radii have historically been the most important
stellar parameters for understanding planetary systems.
With a growing number of planet hosts with precise masses and radii, attention
is turning toward other parameters such as ages to understand the history and
evolution of these systems.
Age is therefore a fundamental stellar parameter of great interest to at least
two large communities of astronomers.
However it is a difficult attribute to measure for main sequence F, G, K, and
M stars in the field, in part because low-mass dwarfs do not move far on the
Color-Magnitude diagram (CMD) during their hydrogen burning lifetimes.
Further, competing stellar evolution models predict different ages for the
same star.
Of the measurable properties for a large ensemble of field stars, rotation
periods contain the most information about stellar age, and provide the best
leverage for advancing our knowledge of galactic archeology as well as
exoplanet population demographics via gyrochronology.

Angular momentum is carried away though magnetically driven stellar winds,
which slows the star’s rotation over cosmic time. This rotation-based `clock'
is known as gyrochronology.
Cool spots on the star’s surface rotate in-to and out-of view, creating small
amplitude ($\pm$ 1\%) quasi-periodic changes in the stellar brightness.
While rotation periods have previously been measured from starspot-induced
flux modulations for hundreds of stars from the ground, space-based
photometric surveys have opened the door to homogeneous ensemble measures of
stellar rotation, and therefore age.
Using precise light curves available from the TESS mission, we hope to
infer ages for nearly 100,000 main sequence field stars, including thousands
of planet hosts.
The ages of the planet hosts will allow us to characterize trends in planet
properties over time, by adding to the \kepler\ sample.

@@

To enable studies of stellar ages from rotation periods with TESS, we propose
to: 1. Measure accurate rotation periods for every available TESS FFI and
two-minute cadence target, using the most appropriate tools and methods
required.
2. Produce updated gyrochronology relations based on a wider range of field
star ages.
3. Determine the origin of the mysterious rotation period bimodality
discovered with Kepler by tracing the rotation period distribution over the
whole sky, and out to further distances utilizing public Gaia data.
4. Measure the star formation history across the sky using a new Bayesian
age-dating system.
5. Further characterize the intrinsic uncertainty in gyrochronology ages using
comoving pairs identified in the first Gaia data release.

@@

the discovery of a bimodal period distribution among field stars by
\citet{mcquillan2013}, which is shown in Figure \ref{davenport}.

@@

The TESS rotation period sample provides the ideal dataset to test these two

@@

this feature as seen in Figure \ref{fig:davenport} with galactic position,
distance, or across the sky.

@@

% Intriguingly, the rotation gap is not constant across masses.
% It appears at shorter rotation periods for more massive stars and longer
% rotation periods for less massive stars.
% In fact, it closely follows a contour of constant age, an isochrone of 600
% Myr.
% Two possible scenarios can explain this phenomenon.
% The first is that stars undergo a transition in the behaviour of their
% magnetic dynamo.
% They switch from a fast to a slow rotating regime rapidly, spending little
% time rotating at an intermittent period.
% This transition would occur around 3-4 days for G dwarfs and 10-12 days for M
% dwarfs.
% Is this a convergence bump?
% The rotation periods of stars are believed to have converged by around the age
% of the Hyades, ~790 Myr (clusters younger than this show a broad spread in
% rotation periods and older clusters a narrow one).
% The age at which this transition occurs, around 600 Myr, is close to this.
% Could this be a coincidence or do rotation periods undergo a rotation period
% bump once they move stabilise onto a Skumanich spin-down regime?
% If this is the case, a study of this transition could greatly advance our
% understanding of the operation of stellar dynamos.

% The rotation period bimodality seen in McQuillan et al. (2014) and Davenport
% (2017) could also be explained by two waves of star formation: a short one
% centered around a few hundred million years ago and a long one centered around
% a few billion years ago.
% TESS’s full view of the sky will allow us to establish the spatial extent of
% this rotation period gap and investigate the underlying cause.

@@

Gyrochronology is by far the most widely applicable, relatively precise
age-dating method available.
Although the phenomenon of magnetic braking has been closely studied since its
discovery in the mid/late 20th century, the manifestations of magnetic fields
and their effects on rotation and activity are complex; the physics
underpinning rotational evolution are far from well-understood.
Although the classical spin-down law of \citet{skumanich1972}: Period
$\propto$ Age$^1/2$ holds for all open clusters with measured periods, it does
not appear to describe old field stars \citep{angus2015, vansaders2016}.
Asteroseismic pulsators observed by Kepler, older than the Sun, rotate more
rapidly than the \citet{skumanich1972} law predicts they should.
\citet{vansaders2016} find that including a transition to a weakened magnetic
braking regime in the gyrochronology models, at a critical Rossby number,
provides an improved fit to the data.
Once a star reaches a critical Rossby number (the ratio of the rotation
period to the convective overturn time), its magnetic dynamo transitions into
a different state which drastically reduces (basically switches off) braking
efficiency.
This means a 5 Gyr star of Solar mass will have the same rotation period as a
6 Gyr and rotation period cannot be used as an age proxy.
The transition occurs at around Solar age for Solar mass stars but later for
less massive stars and earlier for more massive stars.
So, although this effect reduces the usefulness of gyrochronology for stars
more massive than the Sun, in practise, for Solar masses and below,
gyrochronology can still be considered an effective dating method.

Since Solar-type magnetic dynamos are generated through a combination of
convective motion and rotation, gyrochronology is thought to only be
applicable to stars with significant convective envelopes (T$_{\mathrm{eff}}
\lesssim$ 6250 K) that are not yet fully convective (3250 $\lesssim \mathrm{T}
_{\mathrm{eff}}$).
